---
title: "Assignment: End-to-end data analysis in R"
author: "Maria Paula Munoz"
output: html_document
date: "2024-05-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

For this assignment, I've decided to use data from the Shark Control Program in Queensland (https://qfish.fisheries.qld.gov.au/query/number-caught-by-area-calendar-year-and-species-group/table). 

I've already downloaded customized data for the information I need because the initial download contained many blank spaces and NAs. To address this, I cleaned the data on the website using the available functions before downloading.

#### Import data

```{r}
library(readr)
sharkdatacustom <- read_csv("../data/sharkdatacustom.csv")
```

### Making the graphs 

The following chunks show the packages used for this graphs

```{r}
library(tidyverse)
```

For this QFISH analysis I've decided to make two following graph:

1. Time Series Plot: I want to show the trend of the number of catches over the years for different species groups.

#### Code for the graphs 
The following code shows the code I used to create the two graphs mentioned above. It includes specific packages if needed besides the tidyverse, and specific ggplot2 code for the specifics of the graph. 

##### Filtering 
The first step to create my plots is filtering out the rows whit the totals of the catch per group as I do not want them to be graphed because they do not correspond to a specific point in time. Also for the bar plot it would be a double count of the total catches per area. 

```{r}
# Filter out the rows that contain 'Total' in SpeciesGroup or Area
filtered_sharkcontroldata <- sharkdatacustom %>% 
  filter(!grepl("Total", SpeciesGroup) & !grepl("Total", Area))

```

##### Time series plot 

Specific packages needed 
```{r}
library(ggplot2)  # creating plots
library(dplyr)    # data manipulation
```

Time Series Plot code 
```{r}
ggplot(filtered_sharkcontroldata , aes(x = CalendarYear, y = `NumberCaught Total`, color = SpeciesGroup, group = SpeciesGroup)) +
  geom_line() +                             # Add lines connecting the points
  geom_point() +                            # Add points at each data point
  facet_wrap(~Area, scales = "free_y") +    # Create separate plots for each area with independent y scales
  scale_x_continuous(breaks = seq(min(filtered_sharkcontroldata $CalendarYear, na.rm = TRUE), 
                                  max(filtered_sharkcontroldata $CalendarYear, na.rm = TRUE), by = 2)) + # Set x-axis breaks  every 2 years so the years don't overlap
  labs( x = "Year", 
       y = "Number Caught") +               # Add labels 
  theme_minimal() +                         # Use a minimal theme 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) # Rotate and adjust x-axis text                      

# Save the plot with specified dimensions
ggsave("time_series_plot2.png", width = 14, height = 8)
```



 