---
title: "Assignment: End-to-end data analysis in R"
author: "Maria Paula Munoz"
output: html_document
date: "2024-05-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

For this assignment, I've decided to use data from the Shark Control Program in Queensland (https://qfish.fisheries.qld.gov.au/query/number-caught-by-area-calendar-year-and-species-group/table). 

I've already downloaded customized data for the information I need because the initial download contained many blank spaces and NAs. To address this, I cleaned the data on the website using the available functions before downloading.

#### Import data

```{r}
library(readr)
sharkdatacustom <- read_csv("../data/sharkdatacustom.csv")
```

#### Making the graphs 

The following chunks show the packages used for this graphs

```{r}
library(tidyverse)
```

For this QFISH analysis I've decided to make two graphs that are the following:

1. Time Series Plot: I want to show the trend of the number of catches over the years for different species groups.
2. Bar Plot: I want to show the total number of catches per species group per area.

##### Code for the graphs 
The following code shows the code I used to create the two graphs mentioned above. It includes specific packages if needed besides the tidyverse, and specific ggplot2 code for the specifics of the graph. 

The first step to create my plots is filtering the rows whit the totals of the catch per group as I do not want them to be graphed because they do not correspond to a specific point in time. Also for the bar plot it would be a double count of the total catches per area. 

```{r}
# Filter out the rows that contain 'Total' in SpeciesGroup or Area
filtered_data <- data %>% 
  filter(!grepl("Total", SpeciesGroup) & !grepl("Total", Area))

```

```{r}
#TIME SERIES GRAPH

ggplot(filtered_data, aes(x = CalendarYear, y = NumberCaught, color = SpeciesGroup, group = SpeciesGroup)) +
  geom_line() +
  geom_point() +
  facet_wrap(~Area) +
  labs(title = "Number of Catches Over Years by Species Group and Area",
       x = "Year",
       y = "Number Caught") +
  theme_minimal()

```




 