---
title: 'Assigment: Plot deconstruction'
author: "Fernando Cisneros & Maria Paula Muñoz"
date: "2024-05-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##  Plot deconstruction: Plastic Ingestion in Post-hatchling Sea Turtles: Assessing a Major Threat in Florida Near Shore Waters

For this exercise we did a plot deconstruction from a graph picked from a scientific journal. The paper used was from plastic ingestion in sea turtles and we choose a pie chart showing the different types of plastic ingested. 


#### Required packages 

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

```{r}
# Add the the dataset obtained from the supplementary information from the paper
library(readxl)
st <- read_excel("~/Desktop/Data Management in R/Github/MB5370_Mod04_Personal/data/st.xlsx")
#View(st)
```

### Transform to log the total of type plastic ingested 

```{r}
st$log_total <- log(st$total + 1)  # Adding 1 to avoid log(0)

```

```{r}
# Reorder the factor levels of 'type' based on 'total' values in descending order
st$type <- factor(st$type, levels = st$type[order(st$total, decreasing = TRUE)])
```

### Making the new graph from pie chart to barplot. 
The different settings were chosen to add a more scientific look to the graph. 

```{r}
 ggplot(st, aes(x = type, y = log_total)) + 
  geom_bar(stat = "identity", fill = "grey30", color = "black") +  # Set bars to grey with black borders
  xlab("Type of Plastic Ingested") +  # Adding x-axis label
  ylab("Logarithmic Total") +  # Adding y-axis label
  theme_classic() +  # Use a classic theme for a more scientific look
  theme(
    panel.grid.major = element_line(color = "grey80"),  # Major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.text = element_text(size = 12),  # Adjust axis text size
    axis.title = element_text(size = 14, face = "bold"),  # Adjust axis title size and style
    axis.title.x = element_text(margin = margin(t = 10)),  # Add margin to the top of x-axis title
    axis.title.y = element_text(margin = margin(r = 10))   # Add margin to the right of y-axis title
  )
```










